---
title: "885 Project"
author: "Stefan Eng"
date: "`r Sys.Date()`"
engine: knitr
execute:
  echo: false
  message: false
  warning: false
format:
  pdf:
    documentclass: article
    fontsize: 12pt
    geometry:
      - margin=1in
    template-partials:
      - tex_partials/title.tex
      - tex_partials/before-body.tex
include-in-header:
  text: |
    \usepackage{setspace}
    \usepackage{lipsum}
    \doublespacing
    \linespread{2} % Ensures 25 lines per page
    \setlength{\textheight}{9.5in} % Adjust text height for 1-inch margins and 25 lines
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{1em}
abstract: |
  This is the abstract.
---

```{r}
#| echo: false
#| message: false
#| warning: false
library(mgcv)
library(ggplot2)
library(gratia)
library(survival)
library(survminer)
library(ggsurvfit)
library(scales)
library(dplyr)
data(heart)
heart2 <- read.table("Heart_data.txt", skip = 18, header = TRUE)
heart2$reject <- as.factor(heart2$reject)
heart2$surgery <- as.factor(heart2$surgery)
heart2$transplant <- as.factor(heart2$transplant)
heart2$survived <- as.factor(heart2$fustat == 0)
heart$transplant_fct <- as.factor(heart$transplant)
heart2$transplant_fct <- as.factor(heart2$transplant)
heart2$event <- heart2$fustat
levels(heart$transplant_fct) <- c("No transplant", "Transplant")

rm(heart)
heart <- heart2
# Note: It looks like the heart dataset from class is the same cases but heart in survival
# adds controls

# TODO: Swap out for heart with more subjects
#heart$futime <- heart$stop - heart$start
#heart$age <- heart$age + 48

colortheme2 <- c("#FDB44E", "skyblue")
theme_set(theme_classic())
```

## Introduction

TODO: All chatGPT done, need to re-write

Heart transplantation is a critical intervention for patients with end-stage heart failure, offering a potential extension of life and potential for improved quality of life.
The decision for a doctor to advise a patient to undergo a heart transplant is a complex one as it may negatively influence patient survival as well.
This is influenced by various factors, including patient age and clinical status.
Understanding the impact of heart transplantation on patient survival is essential for optimizing treatment strategies and resource allocation.

The Stanford Heart Transplant dataset (Cite: Crawford 77) provides a valuable resource for studying survival outcomes in patients awaiting or undergoing heart transplantation.
This dataset has been analyzed extensively in the literature, with researchers using it to develop and validate survival models, including Cox proportional hazards models.
Derived from a cohort at Stanford University, this dataset includes detailed time-to-event data, allowing for a nuanced analysis of survival probabilities, the effects of treatment, and the role of patient characteristics over time.

In Crawford 77:
- "A quadratic effect of age is suggested" due to stratified analysis on age

In this study, we leverage the Stanford Heart Transplant dataset (from `survival::heart`) to investigate the survival effects of heart transplantation while accounting for patient age and time-dependent factors. Specifically, we aim to: examine the non-linear interaction between age and transplant for survival outcomes for both logistic and survival models.
We will use both local linear kernel methods and smoothing splines to model the time-varying coefficient models and explore the potential impact of transplant rejection, waiting time, and HLA mismatch on patient survival.

## Results

### Descriptive statistics (TODO: Change name)

There are `r nrow(heart)` subjects from the heart transplant dataset, which adds an additional 69 patients to the 103 original patients from the Stanford Heart Transplant dataset.
These additional patients are all controls as they did not receive a transplant and survived or were censored.
There are a total of `r sum(heart$fustat == 0)` deaths and `r sum(heart$fustat == 1)` censored/survival observations in the dataset (`r label_percent()(mean(heart$fustat))`).
See @tbl-1 in the appendix for a summary of the dataset.

### Indication of non-linear interaction

In @fig-age-transplant-interaction, we observe a potential non-linear interaction between age and transplant status on follow-up time.
Those that survived and did received a transplant appear to have a different relationship between age and follow-up time compared to those that did not receive a transplant.
The univariate smooth spline fit only indicates a quadratic relationship in which follow-up time is increased for median age patients for the survived and transplant group.
There is visually a similar pattern for those patients that died and received transplant but the penalized smoothing spline fit is essentially flat.
We also explored the relationship between age and follow-up time for those that received a transplant by pre-transplant bypass surgery status but did not observe any clear patterns See @fig-surgery-transplant-interaction in the appendix.
These results are consistent with the original work done in [@crowley1977] in which they did a stratified analysis on age groups $[8,40), [40,50), [50,65)$ which suggested a quadratic effect of age though this was not statistically significant.


```{r}
#| label: fig-age-transplant-interaction
#| fig-height: 5
#| fig-width: 7
#| fig-caption: "Interaction between age and transplant status on follow-up time"
ggplot(heart2, aes(x = age, y = futime)) +
  facet_grid(rows = vars(transplant),
             cols = vars(survived),
             labeller = label_both) +
  geom_point() +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "bs")) +
  theme_minimal() +
  ylab("Follow-up time") +
  xlab("Age") +
  ylim(c(0, max(heart2$futime) + 10))
```

### Cox Proportional Hazard age-varying transplant model

```{r}
#| label: fig-gam-survival-surgery
#| fig-height: 6
#| fig-width: 8
#| fig-caption: "Predicted survival probabilities from generalized additive age-varying model. "
gams_model <- gam(
    futime ~ surgery + s(age, by = transplant, bs = "bs"),
    family = cox.ph,
    data = heart,
    weights = event
)

gams_model2 <- gam(
    futime ~ s(age, by = transplant, bs = "bs") + s(age, by = surgery, bs = "bs"),
    family = cox.ph,
    data = heart,
    weights = event
)

# summary(gams_model)

k <- 100
pred_data <- expand.grid(
  futime = seq(0, max(heart$futime), length.out = k),
  age = quantile(heart$age, probs = seq(0.1, 0.5, length.out = 4)),
  transplant = c(0,1),
  surgery = c(0,1))

# pred_data$transplant_fct <- as.factor(pred_data$transplant)
levels(pred_data$transplant) <- c("No transplant", "Transplant")

gam_preds <- predict(
  gams_model,
  newdata = data.frame(pred_data), se.fit = TRUE, type = "response")

plot_data <- cbind(pred_data, fit = gam_preds$fit, lower_ci = gam_preds$fit - 1.96 * gam_preds$se.fit, upper_ci = gam_preds$fit + 1.96 * gam_preds$se.fit)

factor_levels <- c("No transplant/No Surgery", "Transplant/No Surgery", "No transplant/Bypass", "Transplant/Bypass")
  
ggplot(plot_data) +
  facet_wrap(~ as.factor(age), ncol = 2, labeller = labeller(
    `as.factor(age)` = function(x) paste("Age:", round(as.numeric(x), 2))
  )) +
  geom_line(
    aes(x = futime, y = fit,
        color = interaction(
          as.factor(transplant),
          as.factor(surgery)
          ),
        linetype = as.factor(surgery))) +
  geom_ribbon(aes(x = futime, ymin = lower_ci, ymax = upper_ci, fill = interaction(
          as.factor(transplant),
          as.factor(surgery)
          )), alpha = 0.2) +
  theme_classic() +
  scale_linetype_manual(
    values = c(1, 2),
    name = "Surgery",
    labels = c("No bypass", "Bypass Surgery")
  ) +
  scale_color_discrete(
    name = "",
    labels = factor_levels
    ) +
  scale_fill_discrete(
    name = "",
    labels = factor_levels
    ) +
  labs(title = "Predicted survival probabilities from GAM",
       x = "Time", y = "Survival probability") +
  ylim(c(0, 1))
```


## Methods

### Logistic Regression Model with varying coefficients via smoothing splines
Let $Y_i$ be the binary outcome variable indicating whether patient $i$ survived or died.
Denote $T_i$ as the transplant status of a patient.
We model the log-odds of survival as a function of patient age and transplant status using a logistic regression varying-coefficient model:

$$
g(\mu_i) = \beta_0 + \theta_1(\text{age}_i)I(T_i = 1) + \theta_2(\text{age}_i)I(T_i = 0)
$$

where $\theta_i$ are smoothing splines of age for patients with and without transplants.


### Local linear kernel smoothing for age-varying treament

$$
\begin{aligned}
g(\mu_i) &= \theta_1(\text{age}_i) + \theta_2(\text{age}_i) T_i
\end{aligned}
$$




- Local linear fitting with varying coefficients (?)
- Something like this chatGPT answer: https://chatgpt.com/c/674f5f0e-d974-8012-a5d5-e386fa864e14
- Pg 41 of the notes good for this

### Smoothing splines
- Age + transplant
- 

## Methods

## Discussion

## References
...

## Appendix

```{r table1}
#| label: tbl-1
#| tbl-cap: "Cohort characteristics of the 103 Stanford Heart Transplant dataset"
library(table1)
heart2$survival <- as.factor(heart2$fustat == 0)
levels(heart2$survival) <- c("Died", "Survived/Censored")
table1(~ futime + age + surgery + transplant + mscore + reject + wait.time + mismatch + hla.a2 | survival, data = heart2)
```

```{r}
#| label: fig-surgery-transplant-interaction
#| fig-height: 5
#| fig-width: 5
heart2 %>% 
  filter(transplant == 1) %>%
ggplot(., aes(x = age, y = futime)) +
  facet_grid(rows = vars(surgery),
             cols = vars(survived),
             labeller = label_both) +
  geom_point() +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "bs")) +
  theme_minimal() +
  ylab("Follow-up time") +
  xlab("Age") +
  ylim(c(0, max(heart2$futime) + 10))
```

### Exploratory plots
