---
title: "Age-varying coefficient analysis of Stanford Transplant Data"
author: "Stefan Eng"
date: "`r Sys.Date()`"
engine: knitr
bibliography: Project-Varying-coefs.bib
execute:
  echo: false
  message: false
  warning: false
format:
  pdf:
    fig-pos: 'H'
    documentclass: article
    fontsize: 12pt
    geometry:
      - margin=1in
    template-partials:
      - tex_partials/title.tex
      - tex_partials/before-body.tex
include-in-header:
  text: |
    \usepackage{setspace}
    \usepackage{lipsum}
    \doublespacing
    \linespread{2} % Enforce 25 lines per page
abstract: |
  We investigate the relationship between age, heart transplantation, and survival outcomes in the Stanford Heart Transplant dataset using spline and kernel varying-coefficient models for both binary survival outcome and time-to-event outcomes.
  We find some evidence for age-varying transplant term and a stronger age-varying bypass surgery term.
  For the kernel based models, we find that a combination of global and local kernel models fits the data better than a purely local kernel model.
  All models indicate that patients around median age have better outcomes than those that are younger or older which is consistent with the original analysis.
github: https://github.com/stefaneng/885-Project
---

\setlength{\abovedisplayskip}{1pt}
\setlength{\belowdisplayskip}{1pt}

```{r setup}
#| echo: false
#| message: false
#| warning: false
library(mgcv)
library(ggplot2)
library(gratia)
library(survival)
library(survminer)
library(ggsurvfit)
library(scales)
library(dplyr)
library(broom)
heart2 <- read.table("Heart_data.txt", skip = 18, header = TRUE)
heart2$reject <- as.factor(heart2$reject)
heart2$surgery <- as.factor(heart2$surgery)
heart2$transplant <- as.factor(heart2$transplant)
heart2$survived <- as.factor(heart2$fustat == 0)
heart$transplant_fct <- as.factor(heart$transplant)
heart2$transplant_fct <- as.factor(heart2$transplant)
heart2$event <- heart2$fustat
levels(heart$transplant_fct) <- c("No transplant", "Transplant")

transplant_factor_levels <- c("No transplant", "Transplant")

rm(heart)
heart <- heart2
heart$y <- as.factor(1 - heart$event)

colortheme2 <- c("#FDB44E", "skyblue")
theme_set(theme_classic())

cv_results_local_kernel <- readRDS("cv_results_local_kernel.rds")
cv_results_partial_local_kernel <- readRDS("cv_results_partial_local_kernel.rds")

cv_bandwidth <- cv_results_partial_local_kernel$bandwidth[which.min(cv_results_partial_local_kernel$deviance)]
```

## Introduction

Heart transplantation is a critical intervention for patients with end-stage heart failure, offering a potential extension of life and potential for improved quality of life.
The success of a heart transplant depends on factors including patient age and clinical status.
Understanding the impact of heart transplantation on patient survival could impact which patients are selected for transplantation and how they are managed post-transplant.

The Stanford Heart Transplant study [@crowley1977] provides a valuable resource for studying survival outcomes in patients awaiting or undergoing heart transplantation.
This dataset has been analyzed extensively in the literature, with researchers using it to develop and validate survival models, including Cox proportional hazards models.
This dataset includes information on patient age, transplant status, and survival outcomes, making it well-suited for investigating the relationship between age and heart transplantation on survival outcomes.
In this study we use the original 103 patient Stanford Heart Transplant dataset to investigate the survival effects of heart transplantation while accounting for patient age and time-dependent factors.
This study has a follow up 173 patients that adds on additional controls, specifically those that did not receive a transplant and did not have bypass surgery and survived during the follow-up period.
We use this extension for the Cox proportional hazards model to ensure that we have sufficient patients for each combination of transplant and bypass surgery status.

We model the non-linear interaction between age and transplant for survival outcomes for both logistic and survival models using varying-coefficient models.
Varying-coefficient models are a flexible class of models that allow for the estimation of non-linear relationships between covariates and outcomes [@hastie1993].
We will use both local linear kernel methods and smoothing splines [@wood2011] to model the time-varying coefficient model for binary survival outcome and time-to-event data.
Our investigation is largely an exploratory analysis which could form the basis for future research on heart transplantation survival.

## Results

There are a total of `r sum(heart$fustat == 1)` deaths and `r sum(heart$fustat == 0)` censored/survival observations in the dataset (`r label_percent()(mean(heart$fustat))`).
See @tbl-1 in the appendix for a summary of the dataset.
In an exploratory analysis, we observe a potential non-linear interaction between age and both transplant status and previous bypass surgery on follow-up time (@fig-age-transplant-interaction).
Those that survived and did received a transplant appear to have a different relationship between age and follow-up time compared to those that did not receive a transplant.
The univariate smooth spline fit seems to indicates a quadratic relationship in which follow-up time is increased for median age patients for the survived and transplant group.
Visually, there is a similar pattern for those patients that died and received transplant but the penalized B-spline fit is essentially flat.
We see a similar pattern for the interaction between age and bypass surgery on follow-up time in @fig-age-surgery-interaction.
We also explored the relationship between age and follow-up time for those that received a transplant by pre-transplant bypass surgery status but did not observe any clear patterns (see @fig-surgery-transplant-interaction in the Appendix).
These results are consistent with the original work done in [@crowley1977] in which they did a stratified analysis on age groups $[8,40), [40,50), [50,65)$ which suggested a quadratic effect of age though this was not statistically significant.

```{r age-transplant-interaction}
#| label: fig-age-transplant-interaction
#| fig-height: 4
#| fig-width: 7
#| fig-cap: "Interaction between age and transplant status on follow-up time"
ggplot(heart2, aes(x = age, y = futime)) +
  facet_grid(rows = vars(transplant),
             cols = vars(survived),
             labeller = labeller(
               survived = function(x) ifelse(x, "Survived", "Died"),
               transplant = function(x) ifelse(x == 1, "Transplant", "No Transplant")
            )) +
  geom_point() +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "bs")) +
  theme_classic() +
  ylab("Follow-up time") +
  xlab("Age") +
  coord_cartesian(ylim = c(0, max(heart2$futime) + 10))
```

```{r age-surgery-interaction}
#| label: fig-age-surgery-interaction
#| fig-height: 4
#| fig-width: 7
#| fig-cap: "Interaction between age and bypass surgery on follow-up time"
ggplot(heart2, aes(x = age, y = futime)) +
  facet_grid(rows = vars(surgery),
             cols = vars(survived),
             labeller = labeller(
               survived = function(x) ifelse(x, "Survived", "Died"),
               surgery = function(x) ifelse(x == 1, "Bypass Surgery", "No bypass")
            )) +
  geom_point() +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "bs")) +
  theme_classic() +
  ylab("Follow-up time") +
  xlab("Age") +
  coord_cartesian(ylim = c(0, max(heart2$futime) + 10))
```

### Logistic regression age-varying models

We fit a logistic regression model to the binary outcome of survival to investigate the relationship between age, transplant status, and survival.
From the exploratory analysis, we fit four logistic additive models models:

1. Three-way interaction between bypass surgery, transplant, and age with no smooth terms
2. Smooth age-varying interaction between bypass surgery and age
3. Separate smooth age-varying interactions between transplant and bypass surgery
4. Smooth age-varying term for bypass surgery-transplant interaction.

We note that the sample size for the dataset is small and the models are more exploratory in nature as the power is quite low.
We compare the models using the Akaike Information Criterion (AIC) and deviance in @tbl-logit-aic-table in the appendix.
The best model selected by AIC is model (2), which has a smooth age-varying interaction between bypass surgery and a constant term for transplant status.
In @fig-gam-spline-glm, we plot the predicted survival probabilities from the best model along the range of ages for patients with and without bypass surgery and transplant status.
We also include the results of the age-varying transplant model as the AIC is within 1 point of the best model.
The model has an age-varying coefficient for bypass surgery which has substantial non-linear components (empirical degrees of freedom = 3.8, p = 0.0167) for the non-surgery group.
The bypass surgery age-varying term is approximately linear (edf = 1.04, p-value = 0.187).
The odds ratio for transplant effect is 0.20 (95% CI: 0.05, 0.75), which indicates a 5 times lower odds of survival for those that received a transplant.
The odds ratio for prior bypass is 0.33 (95% CI: 0.083, 1.30).
Since we have an age-varying coefficient for bypass surgery, the odds ratio for bypass surgery is not interpretable in the usual sense and we need to consider the interaction with age.
For those that received bypass surgery, the survival probability appears lower than those that did not receive bypass surgery but the confidence intervals are quite wide as there are few patients far away from the median age.
We include the individual centered partial effects in @fig-gam-spline-partial in the appendix.

```{r}
#| label: fig-gam-spline-glm
#| fig-height: 4
#| fig-width: 8
#| fig-cap: "Predicted survival probabilities from GAM with smooth age-varying bypass and transplant term. We predict the survival time subset by transplant status and bypass surgery status along the range of ages. The shaded area represents the 95% Bayesian credible intervals."

source("gam_model_glm.R")

k <- 100
pred_data_logit <- expand.grid(
  age = seq(min(heart$age) + 1, max(heart$age) - 1, length.out = 100),
  transplant = c(0,1),
  surgery = c(0,1))

pred_data_logit$transplant <- as.factor(pred_data_logit$transplant)
pred_data_logit$surgery <- as.factor(pred_data_logit$surgery)

gam_preds_logit <- predict(
  final_gam_model_logit$reml,
  newdata = data.frame(pred_data_logit), se.fit = TRUE, type = "response")

gam_preds_logit_mod2 <- predict(
  gam_mods_glm[[3]]$reml,
  newdata = data.frame(pred_data_logit), se.fit = TRUE, type = "response")

plot_data_logit <- cbind(
  pred_data_logit,
  fit = 1 - gam_preds_logit$fit,
  lower_ci = (1 - gam_preds_logit$fit) - 1.96 * gam_preds_logit$se.fit,
  upper_ci = (1 - gam_preds_logit$fit) + 1.96 * gam_preds_logit$se.fit,
  model = "Age-varying surgery")


plot_data_logit <- bind_rows(
  plot_data_logit,
  cbind(
  pred_data_logit,
  fit = 1 - gam_preds_logit_mod2$fit,
  lower_ci = (1 - gam_preds_logit_mod2$fit) - 1.96 * gam_preds_logit_mod2$se.fit,
  upper_ci = (1 - gam_preds_logit_mod2$fit) + 1.96 * gam_preds_logit_mod2$se.fit,
  model = "Age-varying transplant")
)
  
ggplot(plot_data_logit) +
  facet_grid(
    model ~ surgery,
    labeller = labeller(
      surgery = function(x) ifelse(x == 1, "Bypass Surgery", "No bypass")
  )
) +
  geom_line(
    aes(x = age, y = fit,
        color = as.factor(transplant),
        linetype = as.factor(surgery)
          )) +
  geom_ribbon(aes(x = age, ymin = lower_ci, ymax = upper_ci, fill = as.factor(transplant),
          ), alpha = 0.2) +
  theme_classic() +
  scale_linetype_manual(
    name = "",
    labels = c("No Surgery", "Surgery"),
    values = c("solid", "dashed")
  ) +
  scale_color_manual(
    name = "",
    labels = transplant_factor_levels,
    values = colortheme2
  ) +
  scale_fill_manual(
    name = "",
    labels = transplant_factor_levels,
    values = colortheme2
    ) +
  labs(title = "Predicted survival probabilities from GAM logistic regression",
       x = "Age", y = "Survival probability") +
  coord_cartesian(ylim = c(0, 1)) +
  scale_y_continuous(breaks = c(0, 0.5, 1))
```

### Logistic regression local linear kernel model

We also fit a local linear kernel logistic regression model to the binary outcome of survival to investigate the relationship between age, transplant status, and survival.
The first model we fit is a model that is local for the age-varying transplant, and age-varying bypass surgery.
We used a Gaussian kernel function and used 100 target ages between the minimum and maximum age in the dataset.
For bandwidths of 5, 15, and 25, we fit the local linear logistic regression model and plot the results in @fig-local-kernel-glm.
Our leave-out-out cross-validations results indicate that the optimal bandwidth increases monotonically (See @fig-partial-local-kernel-cv-results) but we do not believe this is simply due to bandwidth selection.
The predictions show that the predicted survival probabilities are sensitive to the bandwidth and that the age-varying bypass surgery term appears to have a different relationship with survival than the age-varying transplant term.
The transplant age-varying term quickly appears linear but the bypass surgery term has a more complex relationship with survival.
Based on the results from the smoothing spline, it seems like a lower bandwidth would be more appropriate for the age-varying bypass surgery term and high bandwidth (global logistic regression) is better for the age-varying transplant term.

To remedy this, we fit what we call a partially local kernel logistic regression model in which we first fit a global logistic regression model with age and transplant terms and then fit a local linear kernel model on the Pearson residuals from the global logistic regression model.
As we can see in @fig-partial-local-kernel-cv-results, the partial local model fits better, based on leave-one-out cross-validation, then the local kernel model.
The optimal bandwidth of `r cv_bandwidth` is shown in @fig-partial-local-kernel-glm with $\pm 3$ bandwidth to illustrate how the model fits for different bandwidth.
This model also corresponds better to the GAM models in which the age-varying bypass surgery term is non-linear and the transplant term is linear.

```{r local-kernel-glm}
#| label: fig-local-kernel-glm
#| fig-height: 4
#| fig-width: 8
#| fig-cap: "Predicted survival probabilities from local linear logistic kernel regression for varying bandwidths. "
source("local_kernel_glm.R")

plot_bandwidth <- c(5,10,20,30)
pred_results <- bind_rows(lapply(plot_bandwidth, function(b) {
  # Compute coefficients for each target age
  results <- fit_local_logistic(heart, target_ages, b)

  pred_newdata <- expand.grid(
    age = target_ages,
    transplant = c(0, 1),
    surgery = c(0, 1)
  )

  pred_newdata$pred <-  predict_local_logistic(results, pred_newdata, type = "response")

  pred_newdata$bandwidth <- b
  pred_newdata
}))

ggplot(pred_results) +
  facet_wrap(~ bandwidth,
             labeller = label_both) +
  geom_line(
    aes(
      x = age,
      y = pred,
      color = factor(transplant),
      linetype = factor(surgery))) +
  scale_linetype_manual(
    name = "",
    labels = c("No Surgery", "Surgery"),
    values = c("solid", "dashed")
  ) +  
  scale_color_manual(
    name = "",
    labels = transplant_factor_levels,
    values = colortheme2
  ) +
  scale_fill_manual(
    name = "",
    labels = transplant_factor_levels,
    values = colortheme2
  ) +
  ylab("Probability of survival") +
  xlab("Age")
```

```{r partial-local-kernel-glm}
#| label: fig-partial-local-kernel-glm
#| fig-height: 4
#| fig-width: 8
#| fig-cap: "Predicted survival probabilities from partial local linear logistic kernel regression for bandwidths around optimal bandwidth."
source("local_global_kernel_glm.R")

# Parameters
pred_newdata <- expand.grid(
  age = target_ages,
  transplant = c(0, 1),
  surgery = c(0, 1)
)
  
plot_bandwidth <- c(cv_bandwidth - 3, cv_bandwidth, cv_bandwidth + 3)
pred_results <- bind_rows(lapply(plot_bandwidth, function(b) {
  theta_df <- fit_local_linear(data, target_ages, b)
  
  pred_newdata$pred <- predict_partial_local_logistic(
    global_fit, theta_df, newdata = pred_newdata, type = "response")
  pred_newdata$bandwidth <- b
  pred_newdata
}))

ggplot(
  pred_results,
  aes(x = age, y = pred,
      color = factor(transplant),
      linetype = factor(surgery))) +
  facet_wrap(~ bandwidth,
             labeller = label_both) +
  geom_line() +
  labs(title = "Predicted survival for partial local linear logistic regression", y = "Probability of survival", x = "Age") +
  scale_linetype_manual(
    name = "",
    labels = c("No Surgery", "Surgery"),
    values = c("solid", "dashed")
  ) +
  scale_color_manual(
    name = "",
    labels = transplant_factor_levels,
    values = colortheme2
  ) +
  scale_fill_manual(
    name = "",
    labels = transplant_factor_levels,
    values = colortheme2
  )
```

```{r partial-local-kernel-glm-cv}
#| label: fig-partial-local-kernel-cv-results
#| fig-height: 3
#| fig-width: 6
#| fig-cap: "Cross validation results for partial local linear kernel logistic regression model with Gaussian kernel. The minimum deviance bandwidth for the partial local kernel model is 11.5."

bind_rows(
  bind_cols(cv_results_local_kernel, model = "Local kernel"),
  bind_cols(cv_results_partial_local_kernel, model = "Partial local kernel")
  ) %>%
ggplot(., aes(x = bandwidth, y = deviance)) +
  facet_wrap(~ model, scales = "free") +
  geom_line() +
  geom_point() +
  theme_classic() +
  geom_vline(
    data = . %>% filter(deviance == min(deviance)),
    aes(xintercept = bandwidth),
    linetype = "dashed",
    color = "grey"
  ) +
  labs(title = "Leave-one-out cross-validation error",
       x = "Bandwidth", y = "Cross-validation deviance")
```

### Cox Proportional Hazard Age-Varying Models

We fit Cox proportional hazards models with age-varying coefficients to the time-to-event data to investigate the relationship between age, transplant status, and survival.
For these models we used the extended heart transplant dataset with 172 patients [@therneau2024].
This adds on additional controls such that there are sufficient patients for each combination of transplant and bypass surgery status.
Based on our exploratory analysis, and for agreement with the logistic regression models, we fit the following three models

1. Three-way interaction between bypass surgery, transplant, and age (no splines).
2. A model with a smooth age-varying bypass surgery effect.
3. A model with a smooth age-varying transplant effect.
4. A model with both transplant and bypass surgery age-varying effect.

In the time-to-event case, we found that the best model was the age-varying transplant effect model based on AIC (@tbl-logit-aic-table in appendix).
To assess how the model fits the data we predict the survival time by transplant status and bypass surgery status at ages 39.7, 47.8, 53.0 which correspond to quantiles of 20%, 50% and 80% as shown in @fig-gam-survival-surgery.
Though the AIC indicates that the smoothing splines should be included in the model, the age-varying transplant splines are almost linear as shown in @fig-gam-cox-spline-partial in the appendix, as well as the empirical degrees of freedom of approximately 1.
The parametric coefficients have surgery hazard ratio 0.463 (95% CI: 0.228, 0.938), and transplant hazard ratio 0.734 (95% CI: 0.435, 1.24).
Since transplant is included as a smooth term we must interpret it in combination with the age-varying effect as shown in @fig-gam-survival-surgery.

```{r}
#| label: fig-gam-survival-surgery
#| fig-height: 6
#| fig-width: 8
#| fig-cap: "Predicted survival probabilities from generalized additive age-varying Cox proportional hazard model. The model includes age-varying coefficient for bypass surgery and intercepts for transplant status. We predict the survival time by transplant status and bypass surgery status with three ages 39.7, 47.8, 53.0 which correspond to quantiles of 20%, 50% and 80%. The shaded area represents the 95% Bayesian credible intervals."

source("gam_model_cox.R")

k <- 100
pred_data <- expand.grid(
  futime = seq(0, max(heart2$futime), length.out = k),
  age = quantile(heart2$age, probs = c(0.2, 0.5, 0.8)),
  transplant = c(0,1),
  surgery = c(0,1))

levels(pred_data$transplant) <- c("No transplant", "Transplant")

gam_preds <- predict(
  final_gam_model_cox$reml,
  newdata = data.frame(pred_data), se.fit = TRUE, type = "response")

plot_data <- cbind(
  pred_data,
  fit = gam_preds$fit,
  lower_ci = gam_preds$fit - 1.96 * gam_preds$se.fit,
  upper_ci = gam_preds$fit + 1.96 * gam_preds$se.fit)

transplant_factor_levels <- c("No transplant", "Transplant")
  
ggplot(plot_data) +
  facet_grid(
    as.factor(age) ~ surgery,
    labeller = labeller(
      `as.factor(age)` = function(x) paste("Age:", round(as.numeric(x), 2)),
      surgery = function(x) ifelse(x == 1, "Bypass Surgery", "No bypass")
  )) +
  geom_line(
    aes(x = futime, y = fit,
        color = as.factor(transplant)
          )) +
  geom_ribbon(aes(x = futime, ymin = lower_ci, ymax = upper_ci, fill = as.factor(transplant),
          ), alpha = 0.2) +
  theme_classic() +
  scale_linetype_manual(
    values = c(1, 2),
    name = "Surgery",
    labels = c("No bypass", "Bypass Surgery")
  ) +
  scale_color_manual(
    name = "",
    labels = transplant_factor_levels,
    values = colortheme2
    ) +
  scale_fill_manual(
    name = "",
    labels = transplant_factor_levels,
    values = colortheme2
    ) +
  labs(title = "Predicted survival probabilities from GAM",
       x = "Time", y = "Survival probability") +
  coord_cartesian(ylim = c(0, 1)) +
  scale_y_continuous(breaks = c(0, 0.5, 1))
```

## Methods

### Varying coefficients via smoothing splines

We varying coefficient models with penalized B-splines for the binary outcome of survival as well a Cox proportional hazards model for time-to-event data.
Let $\mu_i$ be the probability that patient $i$ survives.
We model the log-odds of survival as a function of patient age and transplant status using the following logistic regression varying-coefficient models:

$$
\begin{aligned}
\text{logit}(\mu_i) &= \beta_0 + \beta_1 \cdot \text{surgery}_i + \beta_2 \cdot \text{transplant}_i + \beta_3 \cdot \text{age}_i + \beta_4 \cdot (\text{surgery}_i \cdot \text{transplant}_i)  \\
    &\quad\quad +  \beta_5 \cdot (\text{surgery}_i \cdot \text{age}_i) + \beta_6 \cdot (\text{transplant}_i \cdot \text{age}_i) + \beta_7 \cdot (\text{surgery}_i \cdot \text{transplant}_i \cdot \text{age}_i) \\
\text{logit}(\mu_i) &= \beta_0 + \beta_1 \cdot \text{surgery}_i + \beta_2 \cdot \text{transplant}_i + \theta_{\text{surgery}}(\text{age}_i)\cdot \text{surgery}_i \\
\text{logit}(\mu_i) &= \beta_0 + \beta_1 \cdot \text{surgery}_i + \beta_2 \cdot \text{transplant}_i + \theta_{\text{transplant}}(\text{age}_i)\cdot \text{transplant}_i  \\
\text{logit}(\mu_i) &= \beta_0 + \beta_1 \cdot \text{surgery}_i + \beta_2 \cdot \text{transplant}_i\\
&\quad\quad +\theta_{\text{surgery}}(\text{age}_i)\cdot \text{surgery}_i + \theta_{\text{transplant}}(\text{age}_i) \cdot \text{transplant}_i 
\end{aligned}
$$

where $\theta_{surgery}(x)$ and $\theta_{transplant}(x)$ are smoothing B-splines of age for patients.
We include treatment as a fixed effect as smoothing splines are centered so we need to include an intercept to retain the correct interpretation of the treatment effect.
For the censored time-to-event data, we fit one linear Cox proportional hazards models, and the same three varying coefficient models as in the case of the logistic regression model.

$$
\begin{aligned}
\lambda_{linear}(t_i \mid X) &= \lambda_0(t_i) \exp\{ \beta_0 + \beta_1 \cdot \text{surgery}_i + \beta_2 \cdot \text{transplant}_i + \beta_3 \cdot \text{age}_i + \beta_4 \cdot (\text{surgery}_i \cdot \text{transplant}_i) \\
    &\quad\quad\quad +  \beta_5 \cdot (\text{surgery}_i \cdot \text{age}_i) + \beta_6 \cdot (\text{transplant}_i \cdot \text{age}_i) + \beta_7 \cdot (\text{surgery}_i \cdot \text{transplant}_i \cdot \text{age}_i) \} \\
\lambda_{1}(t_i \mid X) &= \lambda_0(t_i) \exp\{
  \beta_0 + \beta_1 \cdot \text{surgery}_i + \beta_2 \cdot \text{transplant}_i +         \theta_{\text{surgery}}(\text{age}_i)\cdot \text{surgery}_i\} \\
\lambda_{2}(t_i \mid X) &= \lambda_0(t_i) \exp\{
  \beta_0 + \beta_1 \cdot \text{surgery}_i + \beta_2 \cdot \text{transplant}_i +     \theta_{\text{transplant}}(\text{age}_i)\cdot \text{transplant}_i \} \\
\lambda_{3}(t_i \mid X) &= \lambda_0(t_i) \exp\{
  \beta_0 + \beta_1 \cdot \text{surgery}_i + \beta_2 \cdot \text{transplant}_i +  \beta_3 \text{surgery}_i \cdot \text{transplant}_i \\
&\quad\quad\quad\quad + \theta_{\text{surgery}}(\text{age}_i)\cdot \text{surgery}_i +  \theta_{\text{transplant}}(\text{age}_i) \cdot \text{transplant}_i \}
\end{aligned}
$$

All smoothing spline models were fit using B-splines with the mgcv package [@wood2011] in R.
The credible intervals are computed from the method described in [@marra2012].
We fit all models using REML when interpreting the smooth coefficients and ML when performing likelihood ratio tests.

### Local linear kernel smoothing for age-varying treament

We also fit a local linear logistic regression model for the binary outcome of survival using a kernel smoothing approach.
Our setup is similar to the varying coefficient model but we use a kernel function to weight the contribution of each patient to the local linear fit.

$$
\begin{aligned}
\text{logit}(\mu_i) &= \theta_1(\text{age}_i) + \theta_2(\text{age}_i) \text{transplant}_i + \theta_3(\text{age}_i) \text{surgery}_i
\end{aligned}
$$

We use a local linear approximations around fixed neighborhood of fixed age points $a_1, \ldots, a_m$.
That is, in the neighborhood of $a$ we approximate $\theta_1(a) + \theta_2(a) \text{transplant}_i + \theta_3(a) \text{surgery}_i$ as a linear interactions between age-transplant and age-surgery.

$$
A = \alpha_{0} + \alpha_{transplant} \text{transplant}_i + \alpha_{surgery} \text{surgery}_i + \alpha_{1} (\text{age}_i - a) \text{transplant}_i + \alpha_{2} (\text{age}_i - a) \text{surgery}_i
$$

where we estimate $\hat{\theta}_1(a) = \hat{\alpha}_{0}$, $\hat{\theta}_2(a) = \hat{\alpha}_{transplant}$, and $\hat{\theta}_3(a) = \hat{\alpha}_{surgery}$ using the estimating equation

$$
\begin{aligned}
\sum_{i=1}^n K_h(\text{age}_i - x) \ell(Y; A) = 0
\end{aligned}
$$

where $\ell(Y; \mathbf{\alpha})$ is the log-likelihood for a Bernoulli random variable and $K_h(x) = K(x/h)/h$ is a kernel function with bandwidth $h$.
We do this by selecting 100 target ages between the minimum and maximum age in the dataset and fit a local linear model at each target age weighted by the Gaussian kernel.
We estimate $\hat{\theta}_1(x) = \hat{\alpha}_{11}$, $\hat{\theta}_2(x) = \hat{\alpha}_{21}$, and $\hat{\theta}_3(x) = \hat{\alpha}_{31}$ at each target age.
The bandwidth $h$ was selected via leave-one out cross validation using deviance as the loss function

$$
\begin{aligned}
D(h) &= -\frac{2}{n} \sum_{i=1}^n [y_i \log(\hat{p}_i) + (1 - y_i) \log(1 - \hat{p}_i))]
\end{aligned}
$$

where $\hat{p}_i$ is the predicted probability of survival for patient $i$.

### Partially local linear kernel smoothing for age-varying treatment

Based on the results from the local linear kernel model, we fit what we call a partially local linear kernel model.
We first fit a global logistic regression model with age and transplant terms and then fits a local linear kernel model on the residuals.
The global model is

$$
\text{logit}(\mu_i) = \beta_0 + \beta_1 \text{transplant}_i + \beta_2 \text{age}_i
$$

Let $r_i$ be the Pearson residuals from the global model, $r_i = \frac{y_i - \hat{\mu}_i}{\sqrt{\text{Var}(y_i)}}$.
We fit a local linear model to the Pearson residuals $r_i = \theta_1(\text{age}_i) + \theta_2(\text{age}_i) \text{surgery}_i$ where we use local linear approximations to estimate $\hat{\theta}_1$ and $\hat{\theta}_2$ with the intercept terms as describe in the previous section.
We then re-scale the estimates for $\hat{\theta}_1$ and $\hat{\theta}_2$ by multiplying by $\sqrt{\text{Var}(y_i)}$.
This allows us to have a global term for transplant and age and a local term for bypass surgery.

## Discussion

Our analysis of the Stanford Heart Transplant dataset suggests that there is a age-varying effect for prior bypass surgery and potentially an age-varying effect for transplant.
Since this dataset has a relatively small sample size it is difficult to include more covariates due to a lack of power.
Another additional analysis that could be performed would be to perform the Cox proportional hazards assumption using local linear estimation for example using a method similar to [@cai2003].
We also acknowledge that the kernel models are missing standard errors and confidence intervals for the predicted probabilities which is a limitation.
Additionally, there are many potentially useful covariates to include for each patient that are both fixed and time-varying coefficients.
For more complex models and covariates, we would need a larger dataset.
Some potential datasets would be the Scientific Registry of Transplant Recipients [@srtr].

\newpage

## References

::: {#refs}
:::

## Appendix

```{r}
#| label: fig-corr-heatmap
#| fig-height: 5
#| fig-width: 8
#| fig-cap: "Correlation between variables from the heart transplant dataset."
numeric_heart2 <- heart2[sapply(heart2, is.numeric) & ! names(heart2) %in% "fustat"]
spearman_cor <- cor(numeric_heart2, method = "spearman", use = "pairwise.complete.obs")
corrplot::corrplot(
  spearman_cor, diag = F, type = "upper",
  method = "number",
  order = "hclust", is.corr = TRUE,
  )
```

```{r}
#| label: fig-km-plot
#| fig-height: 5
#| fig-width: 8
#| fig-cap: "Non-stratified Kaplan-Meier plot for transplant data. Simply performing the transplant does cause an obvious improvement in survival time."
survival_heart <- survival::heart
survfit2(Surv(start, stop, event) ~ transplant, data = survival_heart) %>%
  ggsurvfit() +
  add_confidence_interval() +
  add_risktable() +
  theme_classic() +
  scale_fill_manual(
    name = "",
    labels = c("No transplant", "Transplant"),
    values = colortheme2
  ) +
  scale_color_manual(
    name = "",
    labels = c("No transplant", "Transplant"),
    values = colortheme2
  )
```

```{r table1}
#| label: tbl-1
#| tbl-cap: "Cohort characteristics of the 103 Stanford Heart Transplant dataset"
library(table1)
heart2$survival <- as.factor(heart2$fustat == 0)
levels(heart2$survival) <- c("Died", "Survived/Censored")
table1(~ futime + age + surgery + transplant + mscore + reject + wait.time + mismatch + hla.a2 | survival, data = heart2)
```

```{r}
#| label: fig-gam-spline-partial
#| fig-height: 4
#| fig-width: 8
#| fig-cap: "Partial centered B-spline effects for logistic regression model."

gratia::draw(final_gam_model_logit$reml, scales = "fixed")
```

```{r}
#| label: fig-gam-cox-spline-partial
#| fig-height: 4
#| fig-width: 8
#| fig-cap: "Partial centered B-spline effects for Cox proportional hazard regression model."

gratia::draw(final_gam_model_cox$reml, scales = "fixed")
```

```{r tbl-coxph-aic-table}
#| label: tbl-coxph-aic-table
#| tbl-cap: "AIC and deviance for Cox proportional hazards GAM models"
kableExtra::kable(
  cbind(model = c("linear", "age-surgery smooth", "age-transplant smooth",
                  "age-transplant and age-surgery smooths"), gam_table_aic[, c('aic', 'deviance')]), digits = 1) %>%
  # bold the best model
  kableExtra::row_spec(row = which.min(gam_table_aic$aic), bold = TRUE)
```

```{r tbl-logit-aic-table}
#| label: tbl-logit-aic-table
#| tbl-cap: "AIC and deviance for logistic regression GAM models"
kableExtra::kable(
  cbind(model = c("linear", "age-surgery smooth", "age-transplant smooth",
                  "age-transplant and age-surgery smooths"), gam_table_aic_logit[, c('aic', 'deviance')]), digits = 1) %>%
  # bold the best model
  kableExtra::row_spec(row = which.min(gam_table_aic_logit$aic), bold = TRUE)
```

```{r}
#| label: fig-surgery-transplant-interaction
#| fig-height: 5
#| fig-width: 8
#| fig-cap: "Surgery and transplant appear to have a non-linear interaction with age. Middle aged patients that did not received surgery and survived tended to have longer follow up time than those at extreme ends of age."

heart2 %>% 
  filter(transplant == 1) %>%
ggplot(., aes(x = age, y = futime)) +
  facet_grid(rows = vars(surgery),
             cols = vars(survived),
             labeller = label_both) +
  geom_point() +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "bs")) +
  ylab("Follow-up time") +
  xlab("Age") +
  ylim(c(0, max(heart2$futime) + 10))
```
